---
title: "Inferring circulating COVID-19 cases from reported deaths"
author: "Thibaut Jombart, Sam Abbott, Amy Gimma, Christopher Jarvis, Timothy Russel, Sam Clifford, Sebastian Funk, Hamish Gibbs, Yang Liu, Kevin van Zandvoort, Rosalind Eggo, Adam Kurchaski, CMMID nCov working group, John Edmunds"
date: "`r format(Sys.time(), '%A %d %B %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: zenburn
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_collapse: no
    toc_depth: 4
    toc_float: yes
    css: !expr here::here('css', 'style.css')
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      fig.width = 8,
                      fig.height = 6,
                      dpi = 150,
                      warning = FALSE,
                      message = FALSE,
                      cache = FALSE)
``` 



<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->

# Data preparation {.tabset .tabset-fade .tabset-pills}

## Outline


* **Load scripts**: loads libraries and useful scripts used in the analyses; all
`.R` files contained in `scripts` at the root of the factory are automatically
loaded

* **Load data**: imports datasets, and may contain some *ad hoc* changes to the
data such as specific data cleaning (not used in other reports), new variables
used in the analyses, etc.



## Load packages

```{r libraries}

library(here)
library(reportfactory)
library(incidence)
library(distcrete)
library(epitrix)
library(tidyverse)
library(projections)
library(parallel)

```



## Load scripts

These scripts will load:

* all local scripts, stored as `.R` filesinside `/scripts/`
* all global scripts, i.e. stored outside the factory in `../scripts/`

```{r read_scripts}

rfh_load_scripts()
path_to_simulator <- here::here("rds", "simulate_cases.rds")
simulate_cases <- readRDS(path_to_simulator)

```








<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->

# Model description {.tabset .tabset-fade .tabset-pills}


## Outline

We aim to estimate the number of currently circulating cases on a
given day given a number of deaths reported recently. 

The principle of the estimation is:

1. for each death, draw a likely date of onset from the onset-to-death delay
   distribution; obtain one date of onset per death

2. allocate a batch of $1/CFR$ cases to each date of onset; for each batch,
   simulate epi trajectories using a branching process (Poisson distribution)
   
3. add cases simulated from the different batches

4. repeat steps 1-3 a large number of times (`n_sim`), to reflect uncertainty on
   the actual dates of onsets

5. put all simulations together into a single `projections` object, and derive
   statistics from the simulations


 


## Parameters of the model

This section contain Tim's information on the various parameters. We use these
data to generate distribution, with discretisation when needed.

* **serial interval**: mean of 4.7 days, s.d. of 2.9 days (log normal
  distribution fit); source:
  https://www.medrxiv.org/content/10.1101/2020.02.03.20019497v2.full.pdf.

* **Onset-to-death distribution**: a Gamma(4.726, 0.3151). Source:
  https://www.mdpi.com/2077-0383/9/2/538

* **$R_0$**: somewhere between 1.6 - 4 depending on where the outbreak is and
  which paper you look at. Perhaps we should give a few different options for
  this as an input with a reasonable default value of 2.0? I only say this as it
  seems to be so context-specific. Source:
  https://wellcomeopenresearch.org/articles/5-17

* **CFR**: see that markdown file I sent to you for my estimates. Chris has some
  of his own. All estimates of this are a bit rubbish atm, but better ones are
  incoming (tomorrow!). For now something between 1-2% is pretty reasonable




## Serial interval

`serial_interval` will be a `distcrete` object containing the serial interval
distribution for a discretised log-normal with mean 4.7 and sd 2.9.

```{r serial_interval}

## r_serial_interval() is a simulator for serial interval delays we use a
## log-normal with provided parameters, but cut the tail to ensure no values
## greater than 50 are simulated
serial_interval <- distcrete("lnorm", w = 0, interval = 1,
                             meanlog = log(4.7),
                             sdlog = log(2.9))

## example
plot(0:50, serial_interval$d(0:50),
     type = "h", col = "#5E9281", lwd = 8, lend = 2,
     xlab = "Days from primary to secondary onset",
     ylab = "Probability",
     main = "Serial interval distribution",
     cex.lab = 1.3, cex.main = 1.5)

```



## Onset to death

`r_onset_death` will generate delays from onset to death using a discretised
Gamma with parameters:

* shape: 4.726
* rate: 0.3151


```{r onset_death}

## temporary tweak to check effect of narrower delays
## onset_death <- distcrete("norm", w = 0, interval = 1,
##                          mean = 14,
##                          sd = 1)

r_onset_death <- function(n = 1, min_delay =1, max_delay = 80) {
  ## r_onset_death() will simulate delays from the above distribution, ensuring
  ## that simulated values do not exceed a given maximum;
  ## note: this is total overkill for now
  
  onset_death <- distcrete("gamma", w = 0, interval = 1,
                           shape = 4.726,
                           rate = 0.3151)
  
  out <- onset_death$r(n)
  to_replace <- (out < min_delay) | (out > max_delay)
  while (any(to_replace)) {
    out[to_replace] <- onset_death$r(sum(to_replace))
    to_replace <- out > max_delay
  }  
  out
}


## example
hist(r_onset_death(10000),
     col = "#5E7192", border = "white", nclass = 30,
     xlab = "Days from onset to death",
     main = "Distribution of delay from onset to death",
     cex.lab = 1.3, cex.main = 1.5, prob = TRUE)

```




## Cases per death

The number of cases associated to a given death is determined as the number of trials (cases)
of a Binomial distribution with 1 "success" (deaths) and a probability of CFR. This is implemented
using a Geometric distribution (discretised exponential).

```{r cases_per_death}

hist(1+rgeom(n = 10000, prob = 0.02),
     col = "#A75848", border = "white", nclass = 30,
     xlab = "Number of cases per death",
     main = "Distribution of the numbers of cases per death, CFR = 2%",
     cex.lab = 1.3, cex.main = 1.5)

```








<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->

# Simulation results {.tabset .tabset-fade .tabset-pills}

## Outline

We provide some results for:

* 1 death, with varying CFR (1%, 2%, 3%, 10%) and reproduction numbers (1.5, 2, 3)

* central estimates for data of countries which recently notified deaths: ... to complete


## How many cases for one death?

We combine different values of CFR (1%, 2%, 3%) and reproduction numbers (1.5,
2, 3) and run 200 simulations for each (each simulation generates 50
trajectories for a given combination of onset dates):

```{r sims_one_case}

set.seed(1)
params <- expand.grid(R = c(1.5, 2, 3),
                      cfr = c(0.01, 0.02, 0.03, 0.1))
params
today <- Sys.Date()

## Function to run simulations for the death dates
if(Sys.info()["sysname"] == "Windows"){
  sim_deaths <- function(death_dates, params = NULL, duration) {
    
    cl <- parallel::makeCluster(max(parallel::detectCores()-1,1))

    clusterCall(cl, function() library(here))
    clusterCall(cl, function() library(reportfactory))
    clusterCall(cl, function() library(incidence))
    clusterCall(cl, function() library(distcrete))
    clusterCall(cl, function() library(epitrix))
    clusterCall(cl, function() library(tidyverse))
    clusterCall(cl, function() library(projections))
    clusterCall(cl, function() library(parallel))
    clusterExport(cl, list("death_dates", "params", "duration"), envir = environment())
    clusterExport(cl, list("simulate_cases","merge_add_projections", "merge_projections", "rotate_x","large_txt"), envir = parent.env(environment()))
    
    out <- parallel::parLapply(
      cl,
      seq_len(nrow(params)),
      function(i) simulate_cases(
                      death_dates,
                      n_sim = 200,
                      R = params$R[i],
                      cfr = params$cfr[i], 
                      duration = duration)
    )
    parallel::stopCluster(cl)
    return(out)
  }
} else {
  cores <- 3
  sim_deaths <- function(death_dates, params = NULL) {
    mclapply(
      seq_len(nrow(params)),
      function(i) simulate_cases(
                      death_dates,
                      n_sim = 200,
                      R = params$R[i],
                      cfr = params$cfr[i],
                      duration = duration),
      mc.cores = cores
    )
  }  
}

sims_one_case <- sim_deaths(today, params, duration = 5)

```

We extract quantiles and averages for the different combinations of parameters,
on the day of the death.

```{r results_one_case}

## function to extract results
extract_results <- function(x, date) {
  x <- cumulate(x$projections)
  x <- x[get_dates(x) == date, ] %>%
    as.integer()
  out <- c(mean(x),
           median(x),
           quantile(x, c(.025, .25, .75, .975)))
  names(out) <- c("average",
                  "median",
                  "lower_95",
                  "lower_50",
                  "upper_50",
                  "upper_95"
                  )
  round(out)
}

## Function for get averages and quantiles
get_summaries <- function(sims_cases, date, params = NULL) {
  results_case <- Reduce(rbind, lapply(sims_cases, extract_results, date))
  results_case <- cbind(params, results_case)
  rownames(results_case) <- NULL
  results_case
}


results_one_case <- get_summaries(sims_one_case, today+5, params)

```

```{r, results = "asis"}

## print table
results_one_case %>%
  format(big.mark=" ") %>%
  knitr::kable()

```


## Country Specific analyses

```{r country_cases_data}


cases_path <- here::here("data",
           "bno_cases_data.csv")

df <- read.csv(cases_path, stringsAsFactors = FALSE) %>% 
  mutate(date = as.Date(date, format = "%d/%m/%Y"))

head(df)

```


## France

Run different values of CFR and reproduction number for France. Run 200 simulations for each (each simulation generates 50
trajectories for a given combination of onset dates):

```{r sims_cases_france}

## Function to get dates of death for a country
extract_death_dates <- function(df, country) {
  country_df <- df[df$country == country & df$new_death > 0,]
  death_dates <- rep(country_df$date, country_df$new_death)
  death_dates  
}

fr_death_dates <- extract_death_dates(df, "France")

fr_max_report_date <- df %>% filter(country == "France") %>% .$date %>% max


## Define country specific parameters - based on report 95%CI with data as of 2020-03-02
## No estimates for France so using current estimates from Italy
params_fr <- expand.grid(R = c(2.0, 3.4),
                      cfr = c(0.01, 0.02, 0.03))
  
sims_fr_cases <- sim_deaths(fr_death_dates, params_fr)


fr_max_date <- max(fr_death_dates)
  
## get table
results_fr_cases <- get_summaries(sims_fr_cases, fr_max_date, params_fr)

fr_cases_total <- df %>% filter(country == "France",
                                   date <= fr_max_date) 

fr_reported_cases <- sum(fr_cases_total$new_case)

results_fr_cases$reported <- fr_reported_cases
results_fr_cases$deaths <- length(fr_death_dates)

## print table
results_fr_cases %>%
  format(big.mark=" ") %>%
  knitr::kable()

```

We extract quantiles and averages for the different combinations of parameters,
on the day of the death.


## Italy


Run different values of CFR and reproduction number for Italy. Run 200 simulations for each (each simulation generates 50
trajectories for a given combination of onset dates):


```{r sims_cases_italy, eval=FALSE, include=FALSE}


it_death_dates <- extract_death_dates(df, "Italy")

## Define country specific parameters - based on report 95%CI with data as of 2020-03-02
## Rt estimates have declined since initial outbreak so these numbers will be conservative?
params_it <- expand.grid(R = c(2.0, 2.4),
                      cfr = c(0.01, 0.02, 0.03, 0.1))

sims_it_cases <- sim_deaths(it_death_dates, params_it)

it_max_date <- max(it_death_dates)

results_it_cases <- get_summaries(sims_it_cases, it_max_date, params_it)

## print table
results_it_cases %>%
  format(big.mark=" ") %>%
  knitr::kable()
  
```


## Japan

Run different values of CFR and reproduction number for Japan. Run 200 simulations for each (each simulation generates 50
trajectories for a given combination of onset dates):



```{r sims_cases_japan, eval=FALSE, include=FALSE}

jp_death_dates <- extract_death_dates(df, "Japan")

## Define country specific parameters - based on report 95%CI with data as of 2020-03-02
## Restricted to 1> R estimates due to bug/feature of branching process?
## Rt estimates have declined since initial outbreak so these numbers will be conservative?
params_jp <- expand.grid(R = c(1.00, 1.4),
                      cfr = c(0.01, 0.02, 0.03))

sims_jp_cases <- sim_deaths(jp_death_dates, params_jp)

jp_max_date <- max(jp_death_dates)

results_jp_cases <- get_summaries(sims_jp_cases, jp_max_date, params_jp)

jp_cases_total <- df %>% filter(country == "Japan",
                                   date <= jp_max_date)

jp_reported_cases <- sum(jp_cases_total$new_case)

results_jp_cases$reported <- jp_reported_cases
results_jp_cases$deaths <- length(jp_death_dates)

## print table
results_jp_cases %>%
  format(big.mark=" ") %>%
  knitr::kable()

```


## South Korea

Run different values of CFR and reproduction number for South Korea. Run 200 simulations for each (each simulation generates 50
trajectories for a given combination of onset dates):



```{r sims_cases_south_korea, include=FALSE}

sk_death_dates <- extract_death_dates(df, "South Korea")

## Define country specific parameters - based on report 95%CI with data as of 2020-03-02
## Rt estimates have declined since initial outbreak so these numbers will be conservative?
params_sk <- expand.grid(R = c(1.2, 1.4),
                      cfr = c(0.01, 0.02, 0.03))

sims_sk_cases <- sim_deaths(sk_death_dates, params_sk)

sk_max_date <- max(sk_death_dates)

results_sk_cases <- get_summaries(sims_sk_cases, sk_max_date, params_sk)

sk_cases_total <- df %>% filter(country == "South Korea",
                                   date <= sk_max_date) 

sk_reported_cases <- sum(sk_cases_total$new_case)

results_sk_cases$reported <- sk_reported_cases
results_sk_cases$deaths <- length(sk_death_dates)

## print table
results_sk_cases %>%
  format(big.mark=" ") %>%
  knitr::kable()
  
```

## China

Run different values of CFR and reproduction number for China based on deaths reported up to the 24th of January. Run 200 simulations for each (each simulation generates 50 trajectories for a given combination of onset dates):


```{r sims_cases_china}

ch_death_dates <- extract_death_dates(df, "China")

## Restrict deaths to those from the 24th Jan and previously
ch_death_dates <- ch_death_dates[as.Date(ch_death_dates) <= as.Date("2020-01-21")]

## Define country specific parameters - based on https://doi.org/10.12688/wellcomeopenres.15718.1 (may have Rt on this by end of today)
params_ch <- expand.grid(R = c(2.0, 2.7),
                      cfr = c(0.01, 0.02, 0.03))

sims_ch_cases <- sim_deaths(ch_death_dates, params_ch, duration = 14)

ch_max_date <- max(ch_death_dates)

results_ch_cases <- get_summaries(sims_ch_cases, ch_max_date + 14, params_ch)


ch_cases_total <- df %>% filter(country == "China",
                                   date <= as.Date("2020-01-24") + 14) 

ch_reported_cases <- sum(ch_cases_total$new_case)

results_ch_cases$reported <- ch_reported_cases
results_ch_cases$deaths <- length(ch_death_dates)


## print table
results_ch_cases %>%
  format(big.mark=" ") %>%
  knitr::kable()
```


## Spain

Run different values of CFR and reproduction number for Spain based on deaths reported up to the 24th of January. Run 200 simulations for each (each simulation generates 50 trajectories for a given combination of onset dates):


```{r sims_cases_spain}

sp_death_dates <- extract_death_dates(df, "Spain")
sp_death_dates <- rep(as.Date('2020-03-04'),2) ## There were two but not in data


## Define country specific parameters - based on https://doi.org/10.12688/wellcomeopenres.15718.1 (may have Rt on this by end of today)
params_sp <- expand.grid(R = c(1.3, 3.1),
                      cfr = c(0.01, 0.02, 0.03))

sims_sp_cases <- sim_deaths(sp_death_dates, params_sp)

sp_max_date <- max(sp_death_dates)

results_sp_cases <- get_summaries(sims_sp_cases, sp_max_date, params_sp)

sp_cases_total <- df %>% filter(country == "Spain",
                                   date <= sp_max_date) 

sp_reported_cases <- sum(sp_cases_total$new_case)
sp_reported_cases <- 202 ## Not updated in data yet

results_sp_cases$reported <- sp_reported_cases
results_sp_cases$deaths <- length(sp_death_dates)

## print table
results_sp_cases %>%
  format(big.mark=" ") %>%
  knitr::kable()
```


## Combine countries

```{r sims_cases_countries}


results_sp_cases$country <- "Spain"
results_fr_cases$country <- "France"
results_jp_cases$country <- "Japan"
results_sk_cases$country <- "South Korea"
results_ch_cases$country <- "China"

results_country_cases <- rbind(results_sp_cases, 
                               results_fr_cases,
                               results_jp_cases,
                               results_sk_cases, 
                               results_ch_cases)

## print table
results_country_cases %>%
  format(big.mark=" ") %>%
  knitr::kable()

country_data_path <- here::here('data',
                                'country_cases.csv')


write.csv(results_country_cases, file = country_data_path,
          row.names = F)

```


<!-- ======================================================= --> 
<!-- ======================================================= --> 
<!-- ======================================================= -->

# System information {.tabset .tabset-fade .tabset-pills}

## Outline

The following information documents the system on which the document was
compiled.


## System 

This provides information on the operating system.

```{r system_info}
Sys.info()
```

## R environment

This provides information on the version of R used:

```{r R_session}
R.version
```


## R packages

This provides information on the packages used:

```{r R_pkg}
sessionInfo()
```
