---
title: "Inferring circulating COVID-19 cases from reported deaths"
author: "Thibaut Jombart, Timothy Russel, Sam Abbott, Christopher Jarvis, Amy Gimma, Sam Clifford, Sebastian Funk, Hamish Gibbs, Yang Liu, Rosalind Eggo, CMMID nCov working group, Adam Kurchaski, John Edmunds"
date: "`r format(Sys.time(), '%A %d %B %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: zenburn
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_collapse: no
    toc_depth: 4
    toc_float: yes
    css: !expr here::here('css', 'style.css')
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      fig.width = 8,
                      fig.height = 6,
                      dpi = 150,
                      warning = FALSE,
                      message = FALSE)
```



<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->

# Data preparation {.tabset .tabset-fade .tabset-pills}

## Outline


* **Load scripts**: loads libraries and useful scripts used in the analyses; all
`.R` files contained in `scripts` at the root of the factory are automatically
loaded

* **Load data**: imports datasets, and may contain some *ad hoc* changes to the
data such as specific data cleaning (not used in other reports), new variables
used in the analyses, etc.



## Load packages

```{r libraries}

library(here)
library(reportfactory)
library(incidence)
library(distcrete)
library(epitrix)
library(tidyverse)
library(projections)
library(parallel)

```



## Load scripts

These scripts will load:

* all local scripts, stored as `.R` filesinside `/scripts/`
* all global scripts, i.e. stored outside the factory in `../scripts/`

```{r read_scripts}

rfh_load_scripts()
path_to_simulator <- here("rds", "simulate_cases.rds")
simulate_cases <- readRDS(path_to_simulator)

```








<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->

# Model description {.tabset .tabset-fade .tabset-pills}


## Outline

We aim to estimate the number of currently circulating cases on a
given day given a number of deaths reported recently. 

The principle of the estimation is:

1. for each death, draw a likely date of onset from the onset-to-death delay
   distribution; obtain one date of onset per death

2. allocate a batch of $1/CFR$ cases to each date of onset; for each batch,
   simulate epi trajectories using a branching process (Poisson distribution)
   
3. add cases simulated from the different batches

4. repeat steps 1-3 a large number of times (`n_sim`), to reflect uncertainty on
   the actual dates of onsets

5. put all simulations together into a single `projections` object, and derive
   statistics from the simulations


 


## Parameters of the model

This section contain Tim's information on the various parameters. We use these
data to generate distribution, with discretisation when needed.

* **serial interval**: mean of 4.7 days, s.d. of 2.9 days (log normal
  distribution fit); source:
  https://www.medrxiv.org/content/10.1101/2020.02.03.20019497v2.full.pdf.

* **Onset-to-death distribution**: a Gamma(4.726, 0.3151). Source:
  https://www.mdpi.com/2077-0383/9/2/538

* **$R_0$**: somewhere between 1.6 - 4 depending on where the outbreak is and
  which paper you look at. Perhaps we should give a few different options for
  this as an input with a reasonable default value of 2.0? I only say this as it
  seems to be so context-specific. Source:
  https://wellcomeopenresearch.org/articles/5-17

* **CFR**: see that markdown file I sent to you for my estimates. Chris has some
  of his own. All estimates of this are a bit rubbish atm, but better ones are
  incoming (tomorrow!). For now something between 1-2% is pretty reasonable




## Serial interval

`serial_interval` will be a `distcrete` object containing the serial interval
distribution for a discretised log-normal with mean 4.7 and sd 2.9.

```{r serial_interval}

## r_serial_interval() is a simulator for serial interval delays we use a
## log-normal with provided parameters, but cut the tail to ensure no values
## greater than 50 are simulated
serial_interval <- distcrete("lnorm", w = 0, interval = 1,
                             meanlog = log(4.7),
                             sdlog = log(2.9))

## example
plot(0:50, serial_interval$d(0:50),
     type = "h", col = "#5E9281", lwd = 8, lend = 2,
     xlab = "Days from primary to secondary onset",
     ylab = "Probability",
     main = "Serial interval distribution",
     cex.lab = 1.3, cex.main = 1.5)

```



## Onset to death

`r_onset_death` will generate delays from onset to death using a discretised
Gamma with parameters:

* shape: 4.726
* rate: 0.3151


```{r onset_death}

## temporary tweak to check effect of narrower delays
## onset_death <- distcrete("norm", w = 0, interval = 1,
##                          mean = 14,
##                          sd = 1)

r_onset_death <- function(n = 1, min_delay =1, max_delay = 80) {
  ## r_onset_death() will simulate delays from the above distribution, ensuring
  ## that simulated values do not exceed a given maximum;
  ## note: this is total overkill for now
  
  onset_death <- distcrete("gamma", w = 0, interval = 1,
                           shape = 4.726,
                           rate = 0.3151)
  
  out <- onset_death$r(n)
  to_replace <- (out < min_delay) | (out > max_delay)
  while (any(to_replace)) {
    out[to_replace] <- onset_death$r(sum(to_replace))
    to_replace <- out > max_delay
  }  
  out
}


## example
hist(r_onset_death(10000),
     col = "#5E7192", border = "white", nclass = 30,
     xlab = "Days from onset to death",
     main = "Distribution of delay from onset to death",
     cex.lab = 1.3, cex.main = 1.5, prob = TRUE)

```




## Cases per death

The number of cases associated to a given death is determined as the sample size
of a Binomial distribution with 1 "success" and a probability of CFR. This is
implemented in [Seb Funk's function](https://github.com/sbfnk/bpmodels/blob/6de58a2fe7c24541f488fe58b0a2dfe04f45b58f/R/utils.r#L12) `rbinom_size`, reproduced here:

```{r cases_per_death}

hist(rbinom_size(n = 10000, x = 1, prob = 0.02),
     col = "#A75848", border = "white", nclass = 30,
     xlab = "Number of cases per death",
     main = "Distribution of the numbers of cases per death, CFR = 2%",
     cex.lab = 1.3, cex.main = 1.5)

```








<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->

# Simulation results {.tabset .tabset-fade .tabset-pills}

## Outline

We provide some results for:

* 1 death, with varying CFR (1%, 2%, 3%, 10%) and reproduction numbers (1.5, 2, 3)

* central estimates for data of countries which recently notified deaths: ... to complete


## How many cases for one death?

We combine different values of CFR (1%, 2%, 3%) and reproduction numbers (1.5,
2, 3) and run 200 simulations for each (each simulation generates 50
trajectories for a given combination of onset dates):

```{r sims_one_case}

set.seed(1)
params <- expand.grid(R = c(1.5, 2, 3),
                      cfr = c(0.01, 0.02, 0.03, 0.1))
params
today <- Sys.Date()


sims_one_case <- mclapply(
    seq_len(nrow(params)),
    function(i) simulate_cases(
                    today,
                    n_sim = 200,
                    R = params$R[i],
                    cfr = params$cfr[i]),
    mc.cores = 3
)


```

We extract quantiles and averages for the different combinations of parameters,
on the day of the death.

```{r results_one_case}

## function to extract results
extract_results <- function(x) {
  x <- cumulate(x$projections)
  x <- x[get_dates(x) == today, ] %>%
    as.integer()
  out <- c(mean(x),
           median(x),
           quantile(x, c(.025, .25, .75, .975)))
  names(out) <- c("average",
                  "median",
                  "lower_95",
                  "lower_50",
                  "upper_50",
                  "upper_95"
                  )
  round(out)
}


## get table
results_one_case <- Reduce(rbind, lapply(sims_one_case, extract_results))
results_one_case <- cbind(params, results_one_case)
rownames(results_one_case) <- NULL

```

```{r, results = "asis"}

## print table
results_one_case %>%
  knitr::kable()

```









<!-- ======================================================= --> 
<!-- ======================================================= --> 
<!-- ======================================================= -->

# System information {.tabset .tabset-fade .tabset-pills}

## Outline

The following information documents the system on which the document was
compiled.


## System 

This provides information on the operating system.

```{r system_info}
Sys.info()
```

## R environment

This provides information on the version of R used:

```{r R_session}
R.version
```


## R packages

This provides information on the packages used:

```{r R_pkg}
sessionInfo()
```
