---
title: "Inferring circulating COVID-19 cases from reported deaths"
author: "Thibaut Jombart, Sam Abbott, Amy Gimma, Christopher Jarvis, Timothy Russel, Sam Clifford, Sebastian Funk, Hamish Gibbs, Yang Liu, Kevin van Zandvoort, Rosalind Eggo, Adam Kurchaski, CMMID nCov working group, John Edmunds"
date: "`r format(Sys.time(), '%A %d %B %Y')`"
output:
  html_document:
    code_folding: hide
    highlight: zenburn
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_collapse: no
    toc_depth: 4
    toc_float: yes
    css: !expr here::here('css', 'style.css')
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      fig.width = 8,
                      fig.height = 6,
                      dpi = 150,
                      dev = c("png", "pdf"),
                      fig.path = "figures/",
                      warning = FALSE,
                      message = FALSE)
```



<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->

# Data preparation {.tabset .tabset-fade .tabset-pills}

## Outline


* **Load scripts**: loads libraries and useful scripts used in the analyses; all
`.R` files contained in `scripts` at the root of the factory are automatically
loaded

* **Load data**: imports datasets, and may contain some *ad hoc* changes to the
data such as specific data cleaning (not used in other reports), new variables
used in the analyses, etc.



## Load packages

```{r libraries}

library(here)
library(reportfactory)
library(incidence)
library(distcrete)
library(epitrix)
library(tidyverse)
library(projections)
library(parallel)
library(cowplot)

```



## Load scripts

These scripts will load:

* all local scripts, stored as `.R` filesinside `/scripts/`
* all global scripts, i.e. stored outside the factory in `../scripts/`

```{r read_scripts}

rfh_load_scripts()
path_to_simulator <- here::here("rds", "simulate_cases.rds")
simulate_cases <- readRDS(path_to_simulator)

```










<!-- ======================================================= -->
<!-- ======================================================= -->
<!-- ======================================================= -->

# Simulation results {.tabset .tabset-fade .tabset-pills}

## Outline

We provide some results for:

* an example figure for simulations for CFR 2% and R = 2

* 1 death, with varying CFR (1%, 2%, 3%, 10%) and reproduction numbers (1.5, 2, 3)




## Example figure

We ran analyses with R = 2 and CFR = 2%

```{r example_analyses}

example_results <- simulate_cases(as.Date("2020-03-01"),
                                  R = 2,
                                  cfr = 0.02,
                                  n_sim = 100)
example_results

```


Figure:

```{r figure_1, fig.width = 8, fig.height = 8}

## extract figures
figure_a <- example_results$plot_projections
figure_b <- example_results$plot_projections_cumul

## customisation
figure_a <- figure_a +
  scale_x_date(date_labels = format("%Y-%m-%d")) +
  labs(title = NULL)

figure_b <- figure_b +
  scale_x_date(date_labels = format("%Y-%m-%d")) +
  labs(y = "Total number\nof cases",
       title = NULL)

plot_grid(figure_a,
          figure_b,
          labels = c("A", "B"),
          ncol = 1,
          label_size = 18,
          align = "v")

```



## How many cases for one death?

We combine different values of CFR (1%, 2%, 3%) and reproduction numbers (1.5,
2, 3) and run 200 simulations for each (each simulation generates 50
trajectories for a given combination of onset dates):

```{r sims_one_case}

set.seed(1)
params <- expand.grid(R = c(1.5, 2, 3),
                      cfr = c(0.01, 0.02, 0.03, 0.1))
params
today <- Sys.Date()

if(Sys.info()["sysname"] == "Windows"){
  cores <- 1
} else{
  cores <- 3
}

## Function to run simulations for the death dates 
sim_deaths <- function(death_dates, n_sim = 400) {
  mclapply(
    seq_len(nrow(params)),
    function(i) simulate_cases(
                    death_dates,
                    n_sim = n_sim,
                    R = params$R[i],
                    cfr = params$cfr[i]),
    mc.cores = cores
  )
}  


```

We run simulations for a single death on the 1st March 2020

```{r }

date_death <- as.Date("2020-03-01")
sims_one_case <- sim_deaths(date_death)

```

We extract quantiles and averages for the different combinations of parameters,
on the day of the death.

```{r results_one_case}

## function to extract results
extract_results <- function(x, date) {
  x <- cumulate(x$projections)
  x <- x[get_dates(x) == date, ] %>%
    as.integer()
  out <- c(mean(x),
           median(x),
           quantile(x, c(.025, .25, .75, .975)))
  names(out) <- c("average",
                  "median",
                  "lower_95",
                  "lower_50",
                  "upper_50",
                  "upper_95"
                  )
  round(out)
}

## Function for get averages and quantiles
get_summaries <- function(sims_cases, date) {
  results_case <- Reduce(rbind, lapply(sims_cases, extract_results, date))
  results_case <- cbind(params, results_case)
  rownames(results_case) <- NULL
  results_case
}


results_one_case <- get_summaries(sims_one_case, date_death)

```

```{r, results = "asis"}

## print table
results_one_case %>%
  knitr::kable()

```






<!-- ======================================================= --> 
<!-- ======================================================= --> 
<!-- ======================================================= -->

# System information {.tabset .tabset-fade .tabset-pills}

## Outline

The following information documents the system on which the document was
compiled.


## System 

This provides information on the operating system.

```{r system_info}
Sys.info()
```

## R environment

This provides information on the version of R used:

```{r R_session}
R.version
```


## R packages

This provides information on the packages used:

```{r R_pkg}
sessionInfo()
```
